---
title: "Stat 437 HW4A"
author: "John Salmon (011745357)"
header-includes:
- \usepackage{bbm}
- \usepackage{amssymb}
- \usepackage{amsmath}
- \usepackage{graphicx,float}
- \usepackage{natbib}
output:
  html_document:
    df_print: paged
  pdf_document: default
fontsize: 11pt
editor_options:
  markdown:
    wrap: 72
---

```{r, echo=FALSE, warning=FALSE}
library(knitr)
opts_chunk$set(fig.align="center",tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

# General rule

Please show your work and submit your computer codes in order to get
points. Providing correct answers without supporting details does not
receive full credits. This HW covers

-   Discriminant analysis

For an assignment or project, you DO NOT have to submit your answers or
reports using typesetting software. However, your answers must be well
organized and well legible for grading. Please upload your answers in a
document to the course space. Specifically, if you are not able to knit
a .Rmd/.rmd file into an output file such as a .pdf, .doc, .docx or
.html file that contains your codes, outputs from your codes, your
interpretations on the outputs, and your answers in text (possibly with
math expressions), please organize your codes, their outputs and your
answers in a document in the format given below:

```         
Problem or task or question ... 
Codes ...
Outputs ...
Your interpretations ...
```

It is absolutely not OK to just submit your codes only. This will result
in a considerable loss of points on your assignments or projects.

# Conceptual exercises: III (Discriminant analysis)

\noindent
3. Exercise 2 of Section 4.7 of the Text, which starts with "It was
stated in the text that classifying an observation to the class for
which (4.12) is largest is equivalent to classifying an observation to
the class for which (4.13) is largest. Prove that this is the case."
(Helpful information on how to prove this is contained in the lecture
video on LDA and "LectureNotes5b_notes.pdf".)

Equation 4.12:
$p+k(X) = \frac{\pi_k\frac{1}{\sqrt{2\pi\sigma}}exp(-\frac{1}{2\sigma^2}(x-\mu_k)^2)}{\sum^{K}_{l=1} \pi_l \frac{1}{2\pi\sigma}exp(-\frac{1}{2\sigma^2}(x-\mu_l)^2)}$

Start by taking the numerator of 4.12 because the class that maximizes
$p_k(x)$ is the class that maximizes the numerator:
$\pi_k * exp(-1\frac{1}{2\sigma^2}(x-\mu_k)^2)$

Next take the log to simplify the exponential function:
$\pi_k * exp(-1\frac{1}{2\sigma^2}(x-\mu_k)^2) =  log(\pi_k) - \frac{1}{2\sigma^2}(x-\mu_k)^2$

Expand the quadratic term: $(x - \mu_k)^2 = x^2 - 2x\mu_k + \mu^2_k$

Put together we have
$log(\pi_k) - \frac{1}{2\sigma^2}(x^2 - 2x\mu_k + \mu^2_k) = log(\pi_k) - \frac{x^2}{2\sigma^2}+\frac{x\mu_k}{\sigma^2}-\frac{\mu^2_k}{2\sigma^2}$

The term $\frac{x^2}{2\sigma^2}$ does not depend on k so it is a
constant term that can be dropped. Doing so leaves us with:
$log(\pi_k) + \frac{x\mu_k}{\sigma^2}-\frac{\mu^2_k}{2\sigma^2}$ with
some re-arrangement we get...:

Equation 4.13:
$\delta_k(x) = x * \frac{\mu_k}{\sigma^2} - \frac{\mu^2_k}{2\sigma^2} + log(\pi_k)$

\noindent
4. Exercise 3 of Section 4.7 of the Text, which starts with "This
problem relates to the QDA model, in which the observations within each
class are drawn from a normal distribution with a class specific mean
vector and a class specific covariance matrix. We consider the simple
case where p = 1; i.e. there is only one feature." (Helpful information
on how to prove this is contained in the lecture video on QDA and
"LectureNotes5b_notes.pdf".)

The density of an observation from the kth class drawn from a 1D normal
distribution is

$f_k(x) = \frac{1}{\sqrt{2\pi\sigma^2_k}}exp(-\frac{(x-\mu_k)^2}{2\sigma^2_k})$
each class has a prior probability of $\pi_k$

The Bayes classifier assigns x to the class that maximizes the product
$\pi_kf_k(x) = \pi_k * \frac{1}{\sqrt{2\pi\sigma^2_k}}exp(-\frac{(x-\mu_k)^2}{2\sigma^2_k})$

taking the log to simplify the exp():
$log(\pi_kf_k(x)) = log(\pi_k)-\frac{1}{2}log(2\pi\sigma^2_k)-\frac{(x-\mu_k)^2}{2\sigma^2_k}$

$log(2\pi)$ is a constant so it can be dropped since it doesn't on k:
$log(pi_k) - \frac{1}{2}log(\sigma^2_k)-\frac{(x-\mu_k)^2}{2\sigma^2_k}$

expand the squared term: $(x-\mu_k)^2 = x^2 - 2x\mu_k + \mu^2_k$ so now
its this:
$log(\pi_k) - \frac{1}{2}log(\sigma^2_k) - \frac{1}{2\sigma^2_k}(x^2 - 2x\mu_k + \mu^2_k)$

distribute:
$\delta_k(x) = log(\pi_k)-\frac{1}{2}log(\sigma^2_k)-\frac{x^2}{2\sigma^2_k}+\frac{x\mu_k}{\sigma^2_k}-\frac{\mu^2_k}{2\sigma^2_k}$

When comparing between two classes to decide which gives the higher
discriminant score, the decision rules says we should pick the class
that maximizes $\delta_k(x)$. The coefficient in front of the $x^2$ term
changes for the different classes and the decision boundary equations is
obtained by setting the discriminant functions for two classes equal to
each other: $\delta_k(x)=\delta_l()x$

This gives a quadratic equation because of the differing $x^2$
coefficients making the decision surface quadratic. The discriminant
functions involve quadratic terms, so the resulting bayes classifier
under the QDA model is not linear. Hence the name Quadratic Discriminant
Analysis.

\noindent
5. Exercise 5 of Section 4.7 of the Text, which starts with "We now
examine the differences between LDA and QDA." (Hint: for this question,
you may also use information from Figures 4.9, 4.10 and 4.11 in the
Text.)

\-(a) If the Bayes decision boundary is linear we can expect that the
QDA model may have slightly better (lower) training error because it has
slightly more flexible model parameters that may allow it to overfit to
the training set. However in the test set, with unseen data the LDA
model should perform best because of the correct match of linear
decision boundary and the lower variance because of its fewer parameters
to estimate compared to QDA.

\-(b) If the Bayes decision boundary is non-linear we can expect similar
results. QDA will likely fit best on the training set performance
because it estimates a separate matrix for each class allowing it to
capture the non-linear decision boundary in the training set. In the
test set LDA will have a higher bias but lower variance, because QDA is
flexible it can capture the non-linear boundary but may have extra
variance if the training set did not have many observations.

\-(c)In general as the sample size increases the test prediction
accuracy of QDA should improve compared to LDA. This is because QDA
relies on estimation and more data allows for better estimates of
predictions. Additionally, with smaller sample sizes QDA has to estimate
more parameters which can lead to problems with over fitting.

\noindent
6. Let $Y$ be the random variable for the class label of a random vector
$X \in \mathbb{R}^p$ (where $p$ is the number of features), such that
$Y \in \mathcal{G}=\{1,\ldots, K\}$ and $\Pr(Y=k)=\pi_k$ for Class $k$
with $k \in \mathcal{G}$, where $K \ge 2$ is the number of classes.
Consider the Gaussian mixture model such that the conditional density of
$X$ when it comes from Class $k$ is
$f_k(x) \sim \text{Gaussian}(\mu_k,\Sigma_k)$. Given the training set
$\mathcal{T}$ of $n$ observations $(x_{1},y_{1}),\ldots,(x_{n},y_{n})$
on $(X,Y)$, where $y_i$ is the class label of observation $x_i$, do the
following:

6.1) Provide the MLEs of $\pi_k$, $\mu_k$ and $\Sigma_k$ for each
$k \in \mathcal{G}$ respectively for the case where all $\Sigma_k$'s are
equal and for the case where not all $\Sigma_k$'s are equal. When
$p >n$, is the MLE of $\Sigma_k$ still accurate? If not, recommend a
different estimator for estimating $\Sigma_k$ and provide details on
this estimator.

When all $\Sigma_k$'s are equal: - $\hat{\pi_k} = \frac{n_k}{n}$ -
$\hat{\mu_k} = \frac{1}{n_k}\sum_{y_i=k}x_i$ -
$\hat{\Sigma_k} = \frac{1}{n}\sum^K_{y_i=k}(x_i-\hat{\mu_k})(x_i-\hat{\mu_k})^T$

When all $\Sigma_k$'s different: - $\hat{\pi_k} = \frac{n_k}{n}$ -
$\hat{\mu_k} = \frac{1}{n_k}\sum_{y_i=k}x_i$ -
$\hat{\Sigma_k} = \frac{1}{n_k}\sum^K_{y_i=k}(x_i-\hat{\mu_k})(x_i-\hat{\mu_k})^T$

For cases when $p > n$ the MLE of $\Sigma_k$ will be unstable because
the sample covariance matrix becomes singular and cannot be inverted. To
address this I would suggest using RDA which uses
$\lambda \text{and} \gamma$ to control the regularization of the
co-variances in LDA and QDA.

6.2) Assume $p=2$ and $K=2$ and $k=1$. For the density
$f_k(x) \sim \text{Gaussian}(\mu_k,\Sigma_k)$, what shape do its
contours take, and how does $\Sigma_k$ control the shape of these
contours? How do you check if the conditional density of $X$ given that
it comes from Class $k$ is Gaussian?

The contours in this case will be ellipses and their shape size and
orientation are determined by the covariance matrix. To if the
conditional density of X given class k, we can use the visual methods
that we went over during lecture, or I think there may be some
statistical tests (Shapiro-Wilk?) that could tell you. Additionally tou
can use MLE to estimate the parameters and look at how it fits.

6.3) Is it true that discriminant analysis will perform badly if the
Gaussian assumption is violated? (Hint: for this question, you may also
use the information provided by Figures 4.10 and 4.11 of the Text.) Let
$X=(X_1,\ldots,X_p)^P$, i.e., $X_1$ up to $X_p$ are the feature
variables. Can discriminant analysis be applied to observations of $X$
when some of $X_j,j=1\ldots,p$ is a discrete variable (such as a
categorical variable)? Explain your answer.

When the Gaussian assumption is violated, there is not a guarantee that
the discriminant model will perform poorly, however it is pretty likely
that it will. The model will likely struggle to capture the true
distribution of the data leading to incorrect classification boundaries.
Especially the LDA model as it is more rigid in its assumption that the
covariance matrices are the same across classes.

Discriminant Analysis cannot directly handle categorical features. The
data it takes in needs to be continuous because the decision boundary is
derived by calculating the likelihood of a data point belonging to a
particular class. The categorical likelihood would be estimated by
counting the frequency of each category so it doesn't fit well with the
Gaussian assumption.

6.4) What is a ROC curve, and what is AUC? How is AUC used to gauge the
performance of a classifier? If you apply the same classifier, say, LDA
or QDA under the same Gaussian mixture model, to two data sets that are
independently generated from the same data generating process, i.e.,
that are independently generated from $(X,Y)$ for classification
problems, and obtain two ROC curves, would the two ROC curves be quite
different? Explain your answer. When there are 3 or more classes, are
the codes provided in the lecture notes able to obtain ROC curves and
their AUC's for LDA and QDA?

The ROC curve shows the ability of a classification model to distinguish
between classes at different thresholds. The two axis represent
Sensitivity or true positive rate against the specificity or false
positive rate. As the threshold level changes the balance between these
two changes and is shown by the curve. AUC is the area under the curve
which summarizes the overall performance of the classifier by providing
a value between 0 and 1. 0.5 would be the typical value of a classifier
that just randomly picks classes to assign to. 1 would be a perfect
classifier and 0 would be a bad classifier.

If you were to create an ROC curve for two synthetically generated data
sets from the same distribution, you would likely get 2 curves that are
very similar but may or may not be identical. This is because both
datasets may have some sampling variability which leads to differences
in the points, and that the decision boundaries that are created by
looking at the training data may be slightly different because of
randomness in data sampling.

I suspect that the code provided in the lecture notes wouldn't be
sufficient to generate ROC curves for 3 or more classes. When there are
more than 3 classes the thresholds become more complicated and I'm not
sure how you would display that with a ROC curve.

6.5) Describe the key similarities and differences, respectively,
between LDA and logistic regression. Provide a situation where
discriminant analysis can still be sensibly applied but logistic
regression is not well-defined.

Similarities: - LDA and Logistic Regression are both supervised learning
techniques for classification tasks. - Both use a linear decision
boundary. - Both methods model the probability of a data point belonging
to a class

Differences: - LDA assumes the features are normally distributed,
Logistic Regression does not care about the distribution of the data. -
LDA estimates the class conditional probabilities and uses Bayes rule to
compute the posterior probability, logistic regression models the
probability of a class with the sigmoid logistic function.

A situation where LDA would work but logistic regression may not, is in
a situation where the data set is high dimensional but the number of
observations is low. This could be the case in research of a rare
genetic disease where there are many gene expressions but not that many
people have the disease with the "correct" gene expressions to study. In
this case LDA can be applied because it just estimates the means and
covariance matrices of the classes (has disease vs doesn't have disease)
even with a small sample set, it can still work. But the logistic
regression model may struggle to estimate the coefficients for its
logistic function. It won't be able to find a stable solution because it
will over fit and struggle with the multidisciplinary of the data.

# Applied exercises: III (Discriminant analysis)

\noindent
8. The following is on software commands:

(8.1) What is the main cause of the message "Warning in `lda.default`(x,
grouping, ...): variables are collinear"? What is the main cause of the
message "Error in `qda.default`(x, grouping, ...) : some group is too
small for 'qda'"?

-   The variables are colinear message is shown when one or more
    variable in the data set form linear combinations of others, this
    results in a covariance matrix that in not invertible which the LDA
    calculation relies on to calculate the discriminant function. This
    introduces instability to the calculations where it becomes overly
    sensitive to small adjustments

-   The too small for qda message means that the groups in the predicted
    (response) variable has too few observations. Because QDA relies on
    estimating separate covariance matrices for each group there need to
    be a sufficient number of observations relative to the number of
    predictors to generate accurate estimates. If the groups are too
    small the estimations can fail or become unstable possibly resulting
    in inaccuracy

(8.2) Provide details on the `list` that `predict{MASS}` returns.

-   For the predict.lda() method, the return of the method is a list
    with three components: Class, posterior, and x. The class specifies
    the predicted class as a factor, posterior gives the posterior
    probabilities for the classes, and x gives the scores of the test
    cases.
-   For predict.qda() the list only contains the posterior and class
    components which are the same as lda. The x component is omitted.

(8.3) The arguments `gamma` and `lambda` of `rda{klaR}` are usually
determined by cross-validation. Can they be set manually?

-   Both the lambda and gamma value can be manually set, using the
    regularization parameter. However, I'm not sure why you would do
    that because with out cross validation its not clear what the values
    to set them to should be.

\noindent
9. We will use the human cancer micro-array data that were discussed in
the lectures and are provided by the R library `ElemStatLearn`
(available at
<https://cran.r-project.org/src/contrib/Archive/ElemStatLearn/>). Pick 3
cancer types "MELANOMA", "OVARIAN" and "RENAL", and randomly select the
same set of $60$ genes for each cancer type. Please use `set.seed(123)`
for the whole of this exercise. Your analysis will be based on
observations for these genes and cancer types.

```{r get data, echo=FALSE}
library(ElemStatLearn)

data(nci)
n = dim(nci)[2]
p = dim(nci)[1]

set.seed(123)
random_selection = sample(1:p, size = 60, replace = FALSE)
cancer_type_check = colnames(nci) %in% c("MELANOMA", "OVARIAN", "RENAL")
cancer_selection = which(cancer_type_check == TRUE)
nci1 = nci[random_selection, cancer_selection]
colnames(nci1) = colnames(nci)[cancer_selection]

cancer = data.frame(t(nci1))
cancer$Type = colnames(nci1)
rownames(cancer) = NULL
table(cancer$Type)
```

9.1) Pick 2 features and visualize the observations using the 2
features. Do you think it is hard to classify the observations based on
the amount of overlap among the 3 neighborhoods of observations in each
of the 3 classes? Here "a neighborhood of observations in a class" is a
"open disk that contains the observations in the class".

```{r, 9_1}
library(ggplot2)
ggplot(data = cancer, aes(x = X3, y = X14, colour = Type)) + geom_point()
```

Based on this plot, I would not say it is difficult to classify the
observations optically. Each color represents a cancer type and the
different colored dots are dispersed pretty randomly and there aren't
clear neighborhoods of colored points. Other than the random cluster of
red (melanoma) points that tend to group towards the top right.

9.2) Apply LDA and report the class-wise error rate for each cancer
type.

```{r, 9_2}
library(MASS)
lda.fit = lda(Type~., data = cancer)
lda.predictions = predict(lda.fit, cancer)
true_cancer_types = cancer$Type
predicted_cancer_type = lda.predictions$class

t = table(predicted_cancer_type, true_cancer_types)
print(t)
cat("Total Accuracy:", (sum(diag(t)) / sum(t))* 100, '%')
```

9.3) Use the library `klaR`, and apply regularized discriminant analysis
(RDA) by setting the arguments `gamma` and `lambda` of `rda{klaR}`
manually so that the resulting class-wise error rate for each cancer
type is zero.

```{r, 9_3}
library(klaR)



rda.fit = rda(Type~., data = cancer, gamma = 0.05, lambda = 0.23)
rda.pred = predict(rda.fit, cancer)
t = table(rda.pred$class, cancer$Type)
print(t)
cat("Total Accuracy:", (sum(diag(t)) / sum(t))* 100, '%')
```

9.4) Obtain the estimated covariance matrices from the RDA and visualize
them using the same strategy in Example 3 in "LectureNotes5c_notes.pdf".
What can you say about the degree of dependence among these genes for
each of the three cancer types? (Hint and caution: the class labels
"MELANOMA", "OVARIAN" and "RENAL" will be ordered alphabetically by R.
So, you need to keep track on which estimated covariance matrix is for
which class. Otherwise, you will get wrong visualization.)

```{r, 9_4}
library(reshape2)
hatSigma1 = rda.fit$covariances[,,1]
hatSigma2 = rda.fit$covariances[,,2]
hatSigma3 = rda.fit$covariances[,,3]

m_hs1 = melt(hatSigma1)
m_hs2 = melt(hatSigma2)
m_hs3 = melt(hatSigma3)
All = rbind(m_hs1, m_hs2, m_hs3)
All$Cancer = rep(c("Melanoma", "Ovarian", "Renal"))
All$Cancer = factor(All$Cancer)

cov_plot = ggplot(data = All, aes(x=Var1, y=Var2, fill = value)) + geom_tile() + scale_fill_gradient2(low = "red", high = "blue") + facet_grid(~Cancer) + ggtitle("Estimated Covariance Matrices for Each Cancer Type")

cov_plot
```

It is somewhat difficult to see because of the clutter, but overall I
would say there is less dependence among genes for Ovarian cancer than
the other two. Renal has some color added along the right most columns
and melanoma has more prominent "+" shaped color suggesting there is
more dependence among that cancer type as well.
