---
title: "LDA"
author: "John Salmon"
date: "`r Sys.Date()`"
output: html_document
---

# LDA Implimented

```{r,data}
library(ElemStatLearn) # library containing data
data(nci); n0=dim(nci)[2]; p0=dim(nci)[1] #get dimensions
set.seed(123)
rSel = sample(1:p0, size=50, replace = FALSE)
chk = colnames(nci) %in% c("BREAST", "LEUKEMIA", "COLON")
cSel = which(chk ==TRUE)
ncia = nci[rSel,cSel]
colnames(ncia) = colnames(nci)[cSel]
# transposed subset data as `tmp`
tmp = data.frame(t(ncia)); tmp$Class=colnames(ncia)
rownames(tmp)=NULL; dim(tmp)
table(tmp$Class)
```

```{r, LDA}
set.seed(123)
library(MASS)
lda.fit = lda(Class~.,data=tmp)
lda.pred = predict(lda.fit,tmp)
TrueClassLabel=tmp$Class
LDAEstimtedClassLabel=lda.pred$class
table(LDAEstimtedClassLabel,TrueClassLabel)
```
Ok so if the gaussian distribution requirement is violated, why do we have such a good classification rate?

Well basically as long as the data points are well separated it doesn't really matter what the distribution is.

Support vector machine is a linear classifier that exploits the abiliity to tilt the hyperplane decision boudary based on the classification margin.



```{r, QDA}
set.seed(123)
qda.fit = qda(Class~.,data=tmp)
```
sample group is too small for the estimation of the classwise covariance matrix. To estimate the matricies $n$ must be $\ge$ $p$.

```{r, regularized RDA}
library(klaR)
rda(formula, data, gamma = 0.05, lambda = 0.2, crossval = FALSE, ...)
```
$\gamma$ (gamma) is related to estimating the covariance matrix
$\lambda$ (lambda) is related to how you are calculating the pooled covariance matrix

with regularized findng lambda and gamma you do a 2d grid search and try to minimize the distance between the two generated matricies.

in this case $\gamma$ is very small so the matrix is closer to the identity matrix
the larger the $\lambda$ the less variability between classwise matricies

```{r, regularized RDA}
library(klaR)
rda.fit = rda(Class~.,data=tmp, gamma = 0.05, lambda = 0.2)
rda.pred = predict(rda.fit, tmp) # classify obs in `tmp`
TrueClassLabel=tmp$Class
rDAEstimtedClassLabel=rda.pred$class
table(rDAEstimtedClassLabel,TrueClassLabel)
```
because of the convex hull the lambda and gamma values can be changed slightly with out ruining the perfect classification.

```{r, extract estimated covariance matricies}
# extract estimated covariance matrices
hatSigma1 = rda.fit$covariances[,,1]
hatSigma2 = rda.fit$covariances[,,2]
hatSigma3 = rda.fit$covariances[,,3]
```

```{r, melt est. covariance matrices}
# convert estimated covariance matrices for plotting 
library(reshape2)
melted_hatSigma1 = melt(hatSigma1)
melted_hatSigma2 = melt(hatSigma2)
melted_hatSigma3 = melt(hatSigma3)
EstSigmaAll=rbind(melted_hatSigma1,melted_hatSigma2, melted_hatSigma3)
EstSigmaAll$Cancer=rep(c("BREAST","COLON","LEUKEMIA"), each=nrow(melted_hatSigma1))
EstSigmaAll$Cancer=factor(EstSigmaAll$Cancer)
```

```{r, plot cov matricies}
library(ggplot2)
ggplot(data=EstSigmaAll, aes(x=Var1, y=Var2, fill=value)) + geom_tile() + scale_fill_gradient2(low="blue", high="red") + facet_grid(~Cancer) + xlab("") + ylab("") + ggtitle("Estimated covariance matrices") + theme(plot.title = element_text(hjust = 0.5))
```
