---
title: "Stat437 Lecture 1b Notes"
author: "John Salmon (011745357)"
output: pdf_document
---


```{r include = FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Lecture 1b: Visualization

[link to slides](https://archive.math.wsu.edu/faculty/xchen/stat437/LectureNotes1b.html#/static-visualization)

Visualizing High Dimensional Data is Difficult. Hopf Fibration [link](https://en.wikipedia.org/wiki/Hopf_fibration)

Two Paradigms: 1. Static Visualization: Snapshot of a data set(non-interactive) 1. 
Dynamic Visualization: A sequence of snapshots of data that a user can interact with (interactive)

Static: An example is a fixed generated graph. Dynamic: An example is a dashboard 
with filters or slicers. Or an animation.

#### Static Packages
There are several packages for static visualizations, Such as ggplot2. The Package, 
igraph, creates graphs. ggdendro creates dendrograms(like tree of life) and tree 
diagrams, and ggmap which visualizes dataframes over maps.

#### Dynamic Packages
ploty, for dynamic data frame visualization using dashboards, and gganimate which
creates animations.

### ggplot2

elements involved in a ggplot2 plot
- data, aesthtic mappings, geometric objects
- statistical transformations
- scales coordinate system
- facet

these elements are integrated via the package grammar

- a plot maps data to visual elements via specific grammar
- ggplot2 builds a plot layer by layer

Selecting a proper coordinate system is important for visualizatoin
- cartesian with (x,y) pairs
- polar with position and angle (think boat or plane... traveling north turn left heading 320)

#### example of ggplot2
```{r, ggplot2 example}
library(ggplot2)
#use help (dataset) to get a description of the data
mpg

pl1 = ggplot(data = mpg) + geom_point(mapping = aes(x= displ, y = hwy))

pl1 #display plot
```
- p1 is the handle for the plot (its an image object)
- the geometric object geom_point are the "points" on the graph
the aesthetic mapping mapping = aes maps "displ" and "hwy
 to x-y coordinates using the original scales of measurements in data "mpg"

```{r, another example}
pl2 = ggplot(data = mpg) + geom_line(mapping = aes(x= displ, y = hwy))

pl2
```

This using geom_line instead of geom_point connects each point with a line

The basic structure of a ggplot2 plot command: ggplot(data = <DATA>) + <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))

#### Elementary Visualization
Using the mpg dataset, we will illustrate several instances of elemetary visualization including
- scatter plot
- density plot, histogram
- boxplot
- barplot, pie chart

These 6 plots are all about summary statistics.
Scatter plot: correlation, clustering
Density plot: frequencies
Histogram: frequency for specific events in specific ranges
Boxplot: five point summary (median, first quartile, 3rd quartile, relative min & max)
Barplot: similar to histogram
Pie Chart: Proportion, or frequencies if normalized

#### Back to the mpg dataset:
```{r, back to mpg}
library(ggplot2)
library(dplyr)
mptmp = mpg %>% select(displ, class, drv, hwy)
head(mptmp)

#convert character variables in mpg into factors
mpg = mpg %>% dplyr::mutate_if(is.character, as.factor)
class(mpg$class)

unique(mpg$class)

class(mpg$drv)
unique(mpg$drv)
#the default shape command will not list all the levels

#scatterplot
ggplot(mpg) + geom_point(aes(x = displ, y = hwy)) #same as before, very easy 1 line

#add color aesthetic
p1a = ggplot(mpg) + geom_point(aes(x = displ, y = hwy, color = class))
p1a

#add shape aesthetic
p1b = ggplot(mpg) + geom_point(aes(x = displ, y = hwy, shape = class))
p1b #not that shape by default only can handle 6 values and as such suv is not given a shape. you must manually specify that there are more shapes needed

#fix shapes
p1c = ggplot(mpg) + geom_point(aes(x=displ, y = hwy, shape = class)) + scale_shape_manual(values = 1:length(unique(mpg$class))) #manually specify number of shapes needed
p1c


#add axis labels and title
p1b + xlab("Engine Displacement") + ylab("Highway Milage") + ggtitle("Efficiency") + scale_shape_manual(values = 1:7) +
theme(plot.title = element_text(hjust = 0.5)) #adding this line moves the title to the center of the plot
```
[link to r shapes](https://www.sthda.com/english/wiki/ggplot2-point-shapes)
levels are ordered first numerically, then alphabetically (numeralphabetic? dictionary order)


#### Density Plot
used to:
- visually check the model assumptions
- visually compare a response's behavior under different conditions

when it comes to diagnostics, We don't even need q-q plot if we estimate density properly
```{r, density plot}
mpg1 = mpg %>% filter(class %in% c("compact", "suv", "2seater"))

p2 = ggplot(mpg1, aes(x=hwy, color = class)) + geom_density(linetype = 'dashed') + theme_bw()
#theme_bw() means background white
p2
```

1. how many modes do you see?
3 each of the peaks
2. which density function shows a degree of skewness (right or left)
compact(green) and suv(blue)
3. if you look at the density plots, say something about the average? or can you?
density does not tell you the average.
4. within each class what is the most likely hwy
technically there isnt one because the likelihood of a specific value for continuous variables is 0% but we can say that for example an suv the most likely range is $14 \pm 2$.

$\mu = E(X) = \int_a^b x f(X) dx = x_0 f(x_0) (b-a)$ for some $x_0 \in [a,b]$ by Mean Value Theorem

#### Histogram
Create a histogram for hwy for each of 3 class

```{r, histogram}
p2c = ggplot(mpg1, aes(x=hwy, fill = class)) + geom_histogram(bins = 20, position = "dodge") + theme_bw()
p2c
```

#### Boxplot
Create a boxplot for hwy for each of the 3 class
```{r, boxplot}
p3 = ggplot(mpg1, aes(x=class, y = hwy)) + geom_boxplot() + theme_bw() + stat_summary(fun.y=mean, geom="Point", shape = 23, size = 4)
p3

```

#### Bar Plot
```{r, barplot}
p4 = ggplot(mpg1)+theme_bw() + geom_bar(aes(x=class,y=hwy,fill=drv),stat='identity', position='dodge')
p4
```

For some classes there are not enough observations to construct a bar.
This is an unbalanced design, there is not an equal number of observations for each class.

#### Pie Chart
```{r, piechart}
library(ggplot2) 
library(scales)
# load `scales` package in order to scale data
mpg2 = mpg1 %>% group_by(class) %>% count() %>% ungroup() %>% mutate(percentage=`n`/sum(`n`)) %>% arrange(desc(class))

# create labels using the percentages
mpg2$labels <- scales::percent(mpg2$percentage)

mpg2

#make the pie chart
pie = ggplot(mpg2) + geom_bar(aes(x="", y = percentage, fill=class), stat = "identity", width = 1) + coord_polar("y", start = 0) + theme_void() + geom_text(aes(x=1, y = cumsum(percentage) - percentage/2, label = labels))
pie
```
- group data via class, obtain counts for each class, compute percentages from counts, and arrange classes by their descending percentages
- caution ungroup() is needed after group_by(...) %>% count()
- The best selling chart is the pie chart?

- coord_polar converts bar plot into pie chart
- theme_void() mean "a completely empty theme; otherwise the resulting pie chart does not look nice
- geom_text adds labels to label different parts of a plot at their designated positions
- labels in mpg2 is used to label the slices.