---
title: "lecture 2d & 2e"
author: "John Salmon"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Lecture 2d

- the textbook for stat 435 is quite elementary.
Packages Needed:
- ggmap
- igraph
- ggdendro
- plotly
- gganimate
```{r, lol}
library(dplyr)
library(ggplot2)
library(ggmap)
library(ggdendro)
library(plotly)
library(gganimate)
```

#### Overview

- visualizing data
- visualizing network data
- visualizing tree-like structures
- dynamic visualization

Example tree data, brain artery mappings or brochi (lung pathways) for the brain if your biological features are abnormal, too many blood vessels (tumor) not enough (stroke) then you have a major problem. But how do you store and analyze this network/tree data?

Another example, Kidney shape can tell you whether a person might have kidney disease. This CT scan process is relatively uninvasive as opposed to cutting you open and taking a biopsy. So being able to analyze the 2d image created is very valuable.

```{r, get data}
mydata = read.csv("/Volumes/T7/Stat_437/Lecture/vehicle-accidents.csv")
mydata$State <- as.character(mydata$State)
mydata$MV.Number = as.numeric(mydata$MV.Number)

mydata = mydata %>% filter(State !="Alaska" & State !="Hawaii")
head(mydata %>% select(State,MV.Number))
```

GGMAP is premium to make a map so its left out here

#### igraph

Binary Relationships, such as interactions between two individuals, a disease transmitting between two subjects, and reaction between two chemicals, among subjects or objects can be represented as networks

- each subject is a node or vertex
- each interaction is an edge between associated nodes

There is a cool study/paper/whatever on how a tweet propogates through the internet.
Some also study the connection of our comupters, servers, and server hubs. They want to see which computer network structure is the most secure.

```{r, get data}
nodes = read.csv("/Volumes/T7/Stat_437/Lecture/Dataset1-Media-Example-NODES.csv",header=T, as.is=T)
head(nodes)
links = read.csv("/Volumes/T7/Stat_437/Lecture/Dataset1-Media-Example-EDGES.csv",header=T, as.is=T)
head(links)
```

Network visualization of media sites mentioning each other through hyperlink, reference, and other means.

```{r, make network graph}
library(igraph)
net=graph_from_data_frame(d=links,vertices=nodes,directed=T)
net

# remove multiple edges and loops
net1=simplify(net,remove.multiple=T,remove.loops=T)
# add "vertex.label" and specify label color
plot(net1, edge.arrow.size=.4, edge.color="blue",vertex.label=V(net)$media, vertex.label.color="black")
```
The R Package igraph create graphs as representation of networks.

relationships can be structured differently. It can be one way, predator -> prey, or bi-way media1 <-> media2. 
It is important to consider the structure of you r relationships

The above is a weighted directed graph.

You can interpret the graph by breaking stuff into sub-graphs and counting the number of relationships or connections.

graph theory and small graph world could be interesting.
toplogy of graphs could also be another thing to look into
ontology too maybe? set of all sets...

there are a lot of different things that can be represented by different graphs. binary vs hyper graph

#### Dendrograms

A dendrogram is a diagram representing a tree.
- in hierarchical clustering we can use it to show the arrangement of clusters
- decision making: illustrates decisions to be made under different circumstances
- phylogenetics, the evolutionary relationships can be represented.

Famous model in stats and ML: Regression tree and Boosting
Anyone who can write a paper describing why trees work so well will probably earn an award

```{r dendrogram}
library(ggdendro)
head(USArrests)
```
Data has 3 crimes, Murder Assault, Rape
UrbanPop is percent of populatoin living in urban areas
we are going to classify or cluster based on the euclidean distance between the 4 features.

```{r, make dendro}
hc <- hclust(dist(USArrests), "ave")  # clustering
ggdendrogram(hc, rotate = TRUE) # rotate by 90 degrees
```
This is based on pairwise distance between each observation, EACH OBSERVATION must be in a row otherwise it is useless
start on the furthest right, there are two distinct clusters, move to the left and add more over time increasing the resolution untill you are satisfied.
Its up to us to decide how many states we want to include in each cluster, and the number of clusters.

# Lecture 2e
## Interactiive and Dynamic Visualizations

#### Overview
Interactive visualization allows users to interact with resulting plots
Dynamic visualization present relationships between features as snapshots of observations at different times.

plotly is used for interactive visualizations and gganimate is for dynamic.

Interactive and dynamic visualizations are better presented via html than pdf files.
plotly can also be screenshotted with webshot and phantomjs (javascript(why????(im going to buy a rope and rickety chair if i have to write it)))
```{r webshot}
install.packages("webshot")
webshot::install_phantomjs()
```

brief syntax
plot_ly(data = data.frame()m type = NULL)
type: character strings specifiying the trace type (scatter, bar, box, etc.)
?plotly::plot_ly for more info

#### Example 1 with iris data
```{r ex 1}
library(plotly)
head(iris)
```
blah blah blah google the iris dataset to learn more

```{r, ex1 pt2 i wont regret this scheme later}
#caution ~ is added before a variable name in this library
p = plot_ly(iris, x = ~Petal.Length, y = ~Petal.Width, color = ~Species, type = "scatter") %>% layout(xaxis=list(title="Petal Length"), yaxis=list(title="Petal Width"))
library(shiny) #NOOOOOOOOOO! 시발 아이고
div(p)

#adjust the ledger position (for hw)

```

```{r ex2}
head(midwest %>% select(percollege,state) %>% data.frame())

p1 = plot_ly(midwest, x = ~percollege, color= ~state, type = "box") %>% layout(xaxis=list(title = "Percent of Population with College Education"))
p1
```
```{r, example 2 saving the plot}
library(htmlwidgets)
htmlwidgets::saveWidget(p1, file = "InteractivePlot2.html")
#yay now you have a random file in the folder 
```

#### Visualization with gganimate

```{r, animate}
install.packages("gapminder")
library(gapminder)
head(gapminder)
```
gapminder has information on life expectancy, populaition, gdp per capita, across `r unique(gapmider$year` different years for different countries

life expectancy vs gdp per cap
```{r, plot}
p = ggplot(gapminder, aes(x = gdpPercap, y = lifeExp, size = pop, colour = country)) + geom_point(show.legend = FALSE) + scale_color_viridis_d() + scale_size(range = c(2, 12)) + scale_x_log10()+ labs(x="GDP Per Capita", y = "LIFE expectancy")
p #Static Visualization Nothing Moves
```
```{r, i cant see enough movement!}
library(gganimate)
library(gifski) #visualization as gif in pdf output
p2 = p+transition_time(year)+labs(title = "Year:{frame_time}")
#p2
#why does it make 100 png images instead of displaying as a graph
animate(p2, fps = 10, width = 750, height = 450)
anim_save(p2, "MyAnimation.gif")
```

WE ARE DONE WITH VISUALIZATION!!!!!!!!
