---
title: "Lecture Notes 2c"
author: "John Salmon"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, lol}
library(ggplot2)
library(dplyr)

```
# Lecture 2c

#### geoom_+scale*manual

```{r}
p5 =ggplot(mpg, aes(cty, hwy)) + facet_wrap(~cyl) + theme_bw() + geom_point(aes(shape = drv, color = drv), size = 1.2) + scale_shape_manual(values = c(2,1,4)) + geom_line(aes(linetype = drv), size=0.3) + scale_linetype_manual(values = rep("dotted", 3))

p5

```

#### Combine 2 ggplot2 plots

```{r, combine 2}
p1 = ggplot(mpg, aes(cty, hwy)) + geom_point()
p2 = ggplot(mpg, aes(displ, hwy)) + geom_point()

#combine the two plots
library(gridExtra)
grid.arrange(p1, p2, nrow = 1)
```
#### Concepts not covered
- stat transforms stat_XXX
- figure margin adjustment: Margin
professor says margin adjustment is a complete Pain in the ass so we arent going to cover it. Information on the above can be found in the book :ggplot2: elegant graphs for data analysis" by Hadly Wickham or in various online resources (chatgpt?).
