---
title: "Stat 437 Project 1"
author: 
     - John Salmon (011745357)
     -
header-includes:
   - \usepackage{bbm}
   - \usepackage{amssymb}
   - \usepackage{amsmath}
   - \usepackage{graphicx,float}
   - \usepackage{natbib}
output:
  pdf_document: default
fontsize: 11pt
---

```{r, echo=FALSE, warning=FALSE}
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

# General rule and information
You must show your work in order to get points. Please prepare your report according to the rubrics on projects that are given in the syllabus. In particular, please note that your need to submit codes that would have been used for your data analysis. Your report can be in .doc, .docx, .html or .pdf format. 

The project will assess your skills in	K-means clustering,Hierarchical clustering, Nearest-neighbor classifier, and discriminant analysis for classification, for which visualization techniques you have learnt will be used to illustrate your findings. 

# Data set and its description

Please download the data set "TCGA-PANCAN-HiSeq-801x20531.tar.gz" from the website https://archive.ics.uci.edu/ml/machine-learning-databases/00401/. A brief description of the data set is given at https://archive.ics.uci.edu/ml/datasets/gene+expression+cancer+RNA-Seq. 

You need to decompress the data file since it is a .tar.gz file. Once uncompressed, the data files are "labels.csv" that contains the cancer type for each sample, and "data.csv" that contains the "gene expression profile" (i.e., expression measurements of a set of genes) for each sample. Here each sample is for a subject and is stored in a row of "data.csv". In fact, the data set contains the gene expression profiles for 801 subjects, each with a cancer type, where each gene expression profile contains the gene expressions for the same set of 20531 genes. The cancer types are: "BRCA", "KIRC", "COAD", "LUAD" and "PRAD". In both files "labels.csv" and "data.csv", each row name records which sample a label or observation is for.  


# Task A. Clustering

For this task, you need to apply k-means and hierarchical clustering to cluster observations into their associated cancer types, and report your findings scientifically and professionally. 
Your laptop may not have sufficient computational power to implement k-means and hierarchical clustering on the whole data set, and genes whose expressions are zero for most of the subjects may not be so informative of a cancer type.

Please use `set.seed(123)` for random sampling via the command `sample`, random initialization of `kmeans`, implementing the gap statistic, and any other process where artificial randomization is needed.

(**Task A1**) Complete the following data processing steps:

*  Filter out genes (from "data.csv") whose expressions are zero for at least 300 subjects, and save the filtered data as R object "gexp2".

*  Use the command `sample` to randomly select 1000 genes and their expressions from "gexp2", and save the resulting data as R object "gexp3".

* Use the command `sample` to randomly select 30 samples and their labels from the file "labels.csv", and save them as R object "labels1". For these samples, select the corresponding samples from "gexp3" and save them as R object "gexpProj1". 

* Use the command `scale` to standard the gene expressions for each gene in "gexpProj1", so that they have sample standard deviation 1. Save the standardized data as R object "stdgexpProj1".

```{r Import Data, include=FALSE}
library(dplyr)
set.seed(123)
gene_data = read.csv('data.csv', row.names = 1) #read row names from sample name column
gene_labels = as.data.frame(read.csv('labels.csv', row.names = 1))
```

```{r, Task A1}
set.seed(123)
#filter genes w/ expression zero for at least 300 subjects (aka columns with more than 300 rows containing 0)
num_zeros = colSums(gene_data == 0)
gexp2 = gene_data[, names(num_zeros[num_zeros < 300])]

#sample 1000 genes and their expressions
random_columns = sample(colnames(gexp2), 1000) #don't have to worry about sample column since its the row.name
gexp3 = gexp2 %>% select(all_of(random_columns))

#Sample 30 samples and their labels from the file "labels.csv"
sample_indices = sample(1:dim(gexp3)[1], size = 30, replace = FALSE)
gexpProj1 = gexp3[sample_indices,]
gexpProj1 <- gexpProj1[order(rownames(gexpProj1)), ]

labels1 = gene_labels %>% filter(rownames(gene_labels) %in% rownames(gexpProj1))
labels1 = labels1 %>% mutate_at('Class', as.factor)

#scale the gene expressions
stdgexpProj1 = scale(gexpProj1)
```



(**Task A2**) 

(**Part 1 of Task A2**) Randomly pick 50 genes and their expressions from "stdgexpProj1", and do the following to these expressions: apply the "gap statistic" to estimate the number of clusters, apply K-means clustering with the estimated number of clusters given by the gap statistic, visualize the classification results using techniques given by "LectureNotes3_notes.pdf.pdf", and provide a summary on classification errors. You may use the command `table` and "labels1" to obtain classification errors. Note that the cluster numbering given by `kmeans` will usually be coded as follows:

```
#   Class  label
#     PRAD  5
#     LUAD  4
#     BRCA  1
#     KIRC  3
#     COAD  2
```

When you apply `clusGap`, please use arguments `K.max=10, B=200,iter.max=100`, and when you use `kmeans`, please use arguments `iter.max = 100, nstart=25, algorithm = c("Hartigan-Wong")`.

```{r, Task A2 Part 1}
library(cluster)
library(ggplot2)
set.seed(123)
fiftygenes = sample(colnames(stdgexpProj1), 50)
fiftygeneset = stdgexpProj1[, fiftygenes]

gap = clusGap(fiftygeneset, FUNcluster = kmeans, K.max = 10, B = 200, iter.max = 100)

optimal_k = maxSE(gap$Tab[, "gap"], gap$Tab[, "SE.sim"], method="Tibs2001SEmax")
plot(gap)
cat("Clusgap Computed Optimal K value:", optimal_k, "\n")

k_means = kmeans(fiftygeneset, optimal_k, iter.max = 100, nstart = 25, algorithm = c("Hartigan-Wong"))

vis = data.frame(fiftygeneset)
vis$Type = labels1 %>% filter(rownames(labels1) %in% rownames(fiftygeneset)) %>% pull(Class)
vis$cluster = factor(k_means$cluster)

classification_table = table(vis$Type, vis$cluster)
print(classification_table)

correct_classifications = sum(diag(classification_table))
total_classifications = sum(classification_table)
classification_errors = total_classifications - correct_classifications
error_rate = classification_errors / total_classifications

cat("Number of classification errors:", classification_errors, "\n")
cat("Accuracy:", 1 - error_rate, "\n")


ggplot(vis, aes(x = gene_19110, y = gene_6280)) + xlab("Gene 19110 Expression") + ylab("Gene 6280 Expression") + theme_bw() + geom_point(aes(shape = Type, color = cluster)) + ggtitle("Clustering results using clusGap to estimate cluster number")

```
The classification results from using clusGap to estimate an optimal number of clusters is not very good. ClusGap estimated that there should be only 1 cluster based on the genes from the 30 subjects used for calculation. Since we know what the true number of cluster labels should be, this is obviously incorrect since there are 5 different labels. The k means clustering made 14 out of 30 errors while classifying and had an accuracy of 0.56% which is fairly decent. However given the context of the way this classifier is assigning observations to a single label, this accuracy will vary wildly depending on the distribution of breast cancer occurrences in the data set it is given. This is beacuse it is assigning every set of 50 gene expressions to the same cancer type: BRCA.


(**Part 2 of of Task A2**) Upon implementing `kmeans` with $k$ as the number of clusters, we will obtain the "total within-cluster sum of squares" $W\left(k\right)$ from the output `tot.withinss` of `kmeans`. If we try a sequence of $k=1,2,3,...,10$, then we get $W\left(k\right)$ for
each $k$ between $1$ and $10$. Let us look at the difference
$\Delta_{k}=W\left(  k\right)  -W\left(  k+1\right)$ for $k$ ranging from $1$ to $9$. The $K^{\ast}$ for which
$$
\left\{\Delta_{k}:k<K^{\ast}\right\}  \gg\left\{  \Delta_{k}:k\geq K^{\ast}\right\}
$$
is an estimate of the true number $K$ of clusters in the data, where $\gg$ means "much larger". Apply this method to obtain an estimate of $K$ for the data you created in **Part 1**, and provide a plot of $W\left(k\right)$ against $k$ for each $k$ between $1$ and $10$. Compare this estimate with the estimate obtained in **Part 1** given by the gap statistic, comment on the accuracy of the two estimates, and explain why they are different.

```{r, Task A2 Part 2}
set.seed(123)
n = 10
W = numeric(n)

for (k in 1:n){
  km = kmeans(fiftygeneset, k, iter.max = 100, nstart = 25, algorithm = c("Hartigan-Wong"))
  W[k] = km$tot.withinss
}

delta = numeric(n-1)
for (i in 1:n-1)
  delta[i] = W[i] - W[i + 1]

K_star = which.max(delta) + 1

plot(1:n, W, type = "b", xlab = "K", ylab = "W(k)")

cat("ClusGap estimate", optimal_k, " clusters\n")
cat("K* estimate: ", K_star, " clusters")

k_means_star = kmeans(fiftygeneset, K_star, iter.max = 100, nstart = 25, algorithm = c("Hartigan-Wong"))
vis$cluster = factor(k_means_star$cluster) #overwrite previous cluster labels with new labels from kmeans with kstar estimate
classification_table = table(vis$Type, vis$cluster)
print(classification_table)

correct_classifications = sum(diag(classification_table))
total_classifications = sum(classification_table)
classification_errors = total_classifications - correct_classifications
error_rate = classification_errors / total_classifications

cat("Number of classification errors:", classification_errors, "\n")
cat("Accuracy:", (1 - error_rate), "\n")
```
Both estimates are different from each other, the ClusGap estimate suggests that 1 cluster should be used, while the K-star method predicted 2 clusters. Examining the classification results after using k_means clustering with the $K^*$ method's estimated number of clusters it is clear that while this estimate (2 labels) is closer to the truth (5 cancer types) the overall accuracy of the model has plummeted. With only 16% of the observations being classified correctly this clustering performance is not satisfactory. 

The difference in the estimates likely comes from the two different processes each method is using to generate the estimate. ClusGap uses the gap statistic which compares the total within-cluster variation for different values of K, it searches for the $k$ value where the observed clustering is most distinct from random noise. The $K*$ method looks at the total within-cluster sum of squares as the value of $k$ increases. Then it identifies the point at which the rate of improvement no longer increases, suggesting that increasing the value of k gives diminishing returns. Examining the plot of $w(k) \text{ vs. } k$ it appears that the curve begins to flatten out around the k = 2 value. These two methods operate on different assumptions as ClusGap assumes that a uniform distribution is the base line, so if the data is similar to noise it may just assume 1 cluster is fine. While the $K^*$ method just observes changes in the sum of squares.

Both estimates are inaccurate compared to the truth because its possible the gene data set either does not have clear cluster boundaries for each cancer type, or including 50 genes invokes the curse of dimensionality where the distance between two individual points become much less informative.


(**Part 3 of of Task A2**) Randomly pick 250 genes and their expressions from "stdgexpProj1", and for these expressions, do the analysis in **Part 1** and **Part 2**. Report your findings, compare your findings with those from **Part 1** and **Part 2**; if there are differences between these findings, explain why. Regard using more genes as using more features, does using more features necessarily give more accurate clustering or classification results? 

```{r, Task A2 Part 3}
set.seed(123)
twofiftygenes = sample(colnames(stdgexpProj1), 250)
twofiftygeneset = stdgexpProj1[, twofiftygenes]

#Use Gap Stat to estimate clusters
gap250 = clusGap(twofiftygeneset, FUNcluster = kmeans, K.max = 10, B = 200, iter.max = 100)
optimal_k = maxSE(gap$Tab[, "gap"], gap$Tab[, "SE.sim"], method="Tibs2001SEmax")
cat("Clusgap Computed Optimal K value:", optimal_k, "\n")

k_means = kmeans(twofiftygeneset, optimal_k, iter.max = 100, nstart = 25, algorithm = c("Hartigan-Wong"))

vis = data.frame(twofiftygeneset)
vis$Type = labels1 %>% filter(rownames(labels1) %in% rownames(twofiftygeneset)) %>% pull(Class)
vis$cluster = factor(k_means$cluster)

classification_table = table(vis$Type, vis$cluster)
print(classification_table)

correct_classifications = sum(diag(classification_table))
total_classifications = sum(classification_table)
classification_errors = total_classifications - correct_classifications
error_rate = classification_errors / total_classifications

cat("Number of classification errors:", classification_errors, "\n")
cat("Accuracy:", 1 - error_rate, "\n")

ggplot(vis, aes(x = gene_10523, y = gene_11438)) + xlab("Gene 10523 Expression") + ylab("Gene 11438 Expression") + theme_bw() + geom_point(aes(shape = Type, color = cluster)) + ggtitle("Clustering results using clusGap to estimate cluster number for 250 genes")
```
Increasing the number of genes in the data set does not appear to have an effect on the results if the Gap Statistic is used to estimate the optimal number of clusters. The estimated optimal value for K still remains as 1 and the accuracy and confusion matrix show essentially no change in the results of the classification task. This is likely because Gap Statistic is comparing the clustering results to typical noise, and introducing more features could potentially add more noise to the data. Additionally the curse of dimensionality is further amplified as there are an extra 200 dimensions with distance values separating the points.

```{r, k* with 250 genes}
# Use K* method to estimate optimal number of clusters
set.seed(123)
n = 10
W = numeric(n)

for (k in 1:n){
  km = kmeans(twofiftygeneset, k, iter.max = 100, nstart = 25, algorithm = c("Hartigan-Wong"))
  W[k] = km$tot.withinss
}

delta = numeric(n-1)
for (i in 1:n-1)
  delta[i] = W[i] - W[i + 1]

K_star = which.max(delta) + 1

plot(1:n, W, type = "b", xlab = "K", ylab = "W(k)")

cat("K* estimate: ", K_star, " clusters")

k_means_star = kmeans(twofiftygeneset, K_star, iter.max = 100, nstart = 25, algorithm = c("Hartigan-Wong"))

vis = data.frame(twofiftygeneset)
vis$Type = labels1 %>% filter(rownames(labels1) %in% rownames(twofiftygeneset)) %>% pull(Class)
vis$cluster = factor(k_means_star$cluster) #overwrite previous cluster labels with new labels from kmeans with kstar estimate

classification_table = table(vis$Type, vis$cluster)
print(classification_table)

correct_classifications = sum(diag(classification_table))
total_classifications = sum(classification_table)
classification_errors = total_classifications - correct_classifications
error_rate = classification_errors / total_classifications

cat("Number of classification errors:", classification_errors, "\n")
cat("Accuracy:", (1 - error_rate), "\n")

ggplot(vis, aes(x = gene_10523, y = gene_11438)) + xlab("Gene 10523 Expression") + ylab("Gene 11438 Expression") + theme_bw() + geom_point(aes(shape = Type, color = cluster)) + ggtitle("Clustering results using K* to estimate cluster number for 250 genes")
```
Once again, adding extra factors to cluster with does not appear to have an effect on the estimation of the optimal K value for clustering using the k* method. Much like the Clusgap estimate, it still returned the same value for $k$ to use, suggesting 2 clusters would be the best. Similarly to implimenting ClusGap there was also no change in clustering accuracy for the attempt as the confusion matrix and accuracy suggest no change.

In this case introducing more factors did not have any remarkable changes on the performance of the clustering task. However, this may not be the case for applications in other projects. Adding more features can have an impact on the performance of a clustering task because it is introducing extra information for the clustering algorithm to use. If the information is meaningful it can improve clustering results, if the information is misleading or contradictory it can have no impact or a negative impact on the clustering. Because of the curse of dimesionality adding more features can reduce the information contained in the distance between points. The flip side is the blessing of dimensionality where if the clustering structures are well separated, adding more features can emphasisze the distinctness of the cluster structures.

Although I have limited expertise in genetics, I think it is a fair to make an educated guess and say that for this data set in particular there is not a very clear structure for each of the cancer type classes. This means that regardless of how much information is used during the clustering process, the results will remain fairly poor because there is no clustering structure for the methods to interpret. 
 



(**Task A3**) Randomly pick 250 genes and their expressions from "stdgexpProj1", and for these expressions, do the following: respectively apply hierarchical clustering with average linkage, single linkage, and complete linkage to cluster subjects into groups, and create a dendrogram. For the dendrogram obtained from average linkage, find the height at which cutting the dendrogram gives the same number of groups in "labels1", and comment on the clustering results obtained at this height by comparing them to the truth contained in "labels1".
```{r, task A3}
library(ggdendro)
source("Plotggdendro.r") #Hope its ok to resuse this, I thought it was much easier to use in HW3 than the regular ggdendro function.
set.seed(123)
twofiftygenes_h = sample(colnames(stdgexpProj1), 250)
twofiftygeneset_h = stdgexpProj1[, twofiftygenes_h]


rownames(twofiftygeneset_h) = labels1 %>% filter(rownames(labels1) %in% rownames(twofiftygeneset_h)) %>% pull(Class) 

D = dist(twofiftygeneset_h)
hc_average = hclust(D, method = "average")
hc_single = hclust(D, method = "single")
hc_complete = hclust(D, method = "complete")

cutheight = hc_average$height[length(hc_average$height)-4]
ave_dendro_data = dendro_data_k(hc_average, 5)

plot_ggdendro(ave_dendro_data, direction = "tb", heightReferece = cutheight, expand.y = 0.3) + ggtitle("Clustering with Average Linkage")
ggdendrogram(hc_single) + ggtitle("Hierarchical Clustering with Single Linkage")
ggdendrogram(hc_complete) + ggtitle("HC with complete linkage")

table = table(ave_dendro_data$labels$label, ave_dendro_data$labels$clust)
correct_classifications = sum(diag(table))
total_classifications = sum(table)
classification_errors = total_classifications - correct_classifications
error_rate = classification_errors / total_classifications
print(table)
cat("HC Clustering Accuracy:", 1 - error_rate, "%")
```
Examining the dendrogram created for Hierarchical Clustering using average linkage, we can compare the color of the generated clusters with the true labels which are listed on the bottom. With this comparison in mind, it is fair to say this clustering based on 250 unique genes is not very accurate or conclusive. There are many mismatched colors and labels. Additionally 3 of the estimated clusters light blue, green, and pink all contain 3 or fewer observations yet contain labels that are seen in other clusters. Examining the dendrogram from a human perspective, there does not appear to be a clear pattern between the color (HC Predicted Label) and the true cancer class listed at the bottom. Examining the HC clustering accuracy  


# Task B. Classification

For this task, we will use the same data set you would have downloaded. Please use `set.seed(123)` for random sampling via the command `sample` and any other process where artificial randomization is needed. 

(**Task B1**) After you obtain "labels.csv" and "data.csv", do the following:

*  Filter out genes (from "data.csv") whose expressions are zero for at least 300 subjects, and save the filtered data as R object "gexp2".

*  Use the command `sample` to randomly select 1000 genes and their expressions from "gexp2", and save the resulting data as R object "gexp3". 

*  Pick the samples from "labels.csv" that are for cancer type "LUAD" or "BRCA", and save them as object "labels2". For these samples, pick the corresponding gene expressions from "gexp3" and save them as object "stdgexp2"


(**Taks B2**) The assumptions of linear or quadratic discriminant analysis requires that each observation follows a Gaussian distribution given the class or group membership of the observation, and that each observation follows a Gaussian mixture model. In our settings here, each observation (as a row) within a group would follow a Gaussian with dimensionality equal to the number of genes (i.e., number of entries of the row). So, the more genes whose expressions we use for classification, the higher the dimension of these Gaussian distributions. Nonetheless, you need to check if the Gaussian mixture assumption is satisfied. Note that we only consider two classes "LUAD" and "BRCA", for which the corresponding Gaussian mixture has 2 components and hence has 2 bumps when its density is plotted.

Do the following and report your findings on classification:

* Randomly pick 3 genes and their expressions from "stdgexp2", and save them as object "stdgexp2a".

* Randomly pick 60% of samples from "stdgexp2a", use them as the training set, and use the rest as the test set. You can round down the number of samples in the training set by the command `floor` if it is not an integer.


Build a quadratic discriminant analysis model using the training set, and apply the obtained model to the test set to classify each of its observations. You should code "BRCA" as 0 and "LUAD" as 1. If for an observation the posterior probability of being "BRCA" is predicted by the model to be greater than 0.5, the observation is classified as "BRCA". Report via a 2-by-2 table on the classification errors. Note that the predicted posterior probability given by `qda` is for an observation to belong to class "BRCA".

Before building a quadratic discriminant analysis model, you need to check for highly correlated gene expressions, i.e., you need to check the sample correlations between each pair of columns of the training set. If there are highly correlated gene expressions, the estimated covariance matrix can be close to to being singular, leading to unstable inference. You can remove a column from two columns when their contained expressions have sample correlation greater than 0.9 in absolute value.



(**Taks B3**) Do the following:

* Randomly pick 100 genes and their expressions from "stdgexp2", and save them as object "stdgexp2b".

* Randomly pick 75% of samples from "stdgexp2b", use them as the training set, and use the rest as the test set. You can round down the number of samples in the training set by the command `floor` if it is not an integer.

Then apply quadratic discriminant analysis by following the requirements given in **Taks B2**. Compare classification results you find here with those found in **Taks B2**, and explain on any difference you find between the classification results.


(**Taks B4**) Do the following:

* Randomly pick 100 genes and their expressions from "stdgexp2", and save them as object "stdgexp2b".

* Randomly pick 75% of samples from "stdgexp2b", use them as the training set, and use the rest as the test set. You can round down the number of samples in the training set by the command `floor` if it is not an integer.

Then apply k-nearest-neighbor (k-NN) method with neighborhood size k=3 to the test data to classify each observation in the test set into one of the cancer types. Here, for an observation, if the average of being cancer type "BRCA" is predicted by k-NN to be greater than 0.5, then the observation is classified as being "BRCA". Report via a 2-by-2 table on the classification errors. Compare and comment on the classification results obtained here to those obtain in **Taks B3**. If there is any difference between the classification results, explain why. 




